Sampling requirements to detect ecosystem change

### 2020-06-29 R/manuscript_draft.Rmd
made small updates to text

Number of references: 35<!--\total{citenum}-->

@@

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], showing that if a significant trend is observed from a dataset with limited temporal perspective, it is likely to reliably describe qualitatively the complete trend (increase or decline), but is unlikely to provide an accurate quantitative change in population. Other research has investigated the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the benefit of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

@@

Here, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed. We tailor our analysis to the detection of a changepoint in a time series, but our approach is applicable to other questions as well. Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes [@James2014]. We focus on paleoecological core samples as one example of this type of data. We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, i.e. a changepoint. However, our approaches are widely applicable to other questions and data types. We first investigate these tools using a simulation-based approach. We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy [@Bruel2018].

@@

Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019e]. We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series. Monitoring programs have to be able to address our questions of interest with sufficient statistical power. In addition, optimizing sampling efforts is valuable given the high costs of many monitoring programs [@Caughlan2001a;@Bennett2014a]. Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b]. Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future. If only a subsample of the samples can be analyzed, it may be better to choose samples carefully as opposed to random or regular sampling. This can improve the accuracy of the results and reduce costs overall.

# Code availability

>>>>

Number of references: 38<!--\total{citenum}-->

@@

Accounting for key considerations in the design of a monitoring program is necessary to detect long-term ecological change. The specifics of the monitoring program will determine the power with which a question of interest can be addressed. For example, @White2019e found that 72\% of vertebrate populations required at least 10 years of monitoring to detect significant changes in the population size over time. The specific number of years required depended on the species biology and the detection method used [@White2019e]. Other work has focused on the frequency of monitoring [@Wauchope2019a], the impact of allocating monitoring resources spatially versus temporally [@Rhodes2011a; @Weiser2019a], and the costs and benefits of increasing sampling breadth relying on citizen science [@Weiser2020a]. Lastly, both the ecological and economical costs of failing to detect a true trend (type II error) have to be weighed against the risks of false (type I error) detection [@Mapstone1995a]. Given limited budgets, monitoring programs need to be designed to be cost-effective [@Caughlan2001a; @Grantham2008b; @Bennett2016a].

Because ecological systems are dynamic in both space and time, it is essential that sampling designs account for spatio-temporal dynamics [@Williams2018]. To estimate a particular parameter, e.g. population abundance, optimal spatial sampling strategies are based on spatially balanced sampling [@Kermorvant2019], while optimal temporal sampling strategies are more cost efficient with a targeted sampling, e.g., around the period of reproduction [@Jackson2008a]. In either case, sampling time and lcoations can be chosen in an iterative process to be cost-effective and reduce the uncertainty in the process [@Hooten2009]. With the ability to choose precisely when to sample, we can move beyond random, interval, or opportunistic sampling designs. This is particuraly relevant in situations where a subset of samples already collected need to be analyzed.

@@

Building on ideas from dynamic survey design and optimal monitoring research, we develop a set of tools to determine the appropriate number of samples and sampling approach when dealing with data sources where only a subset of samples are analyzed. We tailor our analysis to the detection of a changepoint in a time series, but our approach is applicable to other questions as well. Changepoints are an important characteristic of a time series as they can indicate a underlying change in ecosystem processes [@James2014]. We focus on paleoecological core samples as one example of this type of data. We examine the situation where the goal is to detect the time at which a significant change in an ecological community occurs, i.e. a changepoint. However, our approaches are widely applicable to other questions and data types. We first investigate these tools using a simulation-based approach. We then test the tools on a case study from a paleosequence of Cladocera community assemblage from Lake Varese located in the subalpine region of north-western Italy [@Bruel2018].

@@

Analyzing a subsample of a time series as opposed to the whole time series will inevitably leads to a lesser understanding of the phenomenon observed [@White2019e]. We show here that an informed subsampling can still allow detection of critical information, such as a changepoint in a time series. Monitoring programs have to be able to address our questions of interest with sufficient statistical power. In addition, optimizing sampling efforts is valuable given the high costs of many monitoring programs [@Caughlan2001a;@Bennett2014a]. Thus, costs of monitoring have to weighed against the value gained from monitoring---a value of information approach [@Lovett2007;@Bennett2018b]. Monitoring programs should try to anticipate the potential questions of tomorrow, and reducing the data collected, or analyzed, must be done with the best foresight possible on how these data may be necessary to manage ecosystems in the future. If only a subsample of the samples can be analyzed, it may be better to choose samples strategically as opposed to random or regular sampling. This can improve the accuracy of the results and reduce costs overall.

# Author contributions

ERW and RB conceived the ideas and designed methodology; RB collected the data; ERW designed the simulations; ERW and RB analysed the data; Both authors contributed critically to the drafts and gave final approval for publication

# Data availability
